---
title: "BST 270 Individual Project"
subtitle: "Reproducible Data Science: Joe Biden's 2019 Media Coverage (FiveThirtyEight)"
author: "Tzu-Hsi Jen"
date: "2025-01-20"
format: 
  html: default
  pdf: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Introduction

The purpose of this notebook is to meet the individual project criteria for the course BST270: Reproducible Data Science, taken Winter 2025.

### Motivation and Reproducibility

The aim of this project is to accurately reproduce one figure and one table from FiveThirtyEight's [The Media Frenzy Around Biden Is Fading](https://fivethirtyeight.com/features/the-media-frenzy-around-biden-is-fading/). It utilises datasets capturing media mentions from both online and cable news sources. The online news data, drawn from the Media Cloud database, includes U.S. Top Online News and U.S. Top Digital Native News, and is stored at `../data/online_weekly.csv`. Cable news data, collected through the GDELT Television API from the TV News Archive, can be found at `../data/cable_weekly.csv`.

### Set Up

#### Install Required Packages and Load Libraries

Firts, we install and load the necessary packages for data manipulation and visualisation. This includes `lubridate` for date conversions, `dplyr` and `tidyr` for data organization, `ggplot2` for static plotting, and `plotly` for interactive visualizations. We also use `knitr` for table generating.

```{r}
#install.packages("dplyr")
#install.packages("lubridate")
#install.packages("ggplot2")
#install.packages("tidyr")
#install.packages("plotly")
#install.packages("knitr")

# For data wrangling purposes:
library(lubridate)
library(dplyr)
library(tidyr)

# For generating the plot and the table
library(ggplot2)
library(plotly)
library(knitr)
```

#### Import Data
Next, we import the datasets for analysis.

```{r}
online_data <- read.csv("../data/online_weekly.csv") # online news weekly data
cable_data <- read.csv("../data/cable_weekly.csv") # cable news weekly data
```

### Data Wrangling: Online News

We start by processing `online_data` to compute the weekly percentage of online news mentions for each candidate starting from 2019-05-01. This start date aligns with the timeline presented in Figure 1 of the original article.

```{r}
online_news <- online_data|>
  mutate(date = ymd(date))|> # convert date into ymd format
  filter(date >= "2019-05-01")|> # select data starting from 2019-05-01 as shown in the original figure 
  mutate(week = floor_date(date, unit = "week", week_start = 7))|> # week starts from Sunday
  arrange()|>
  group_by(name, week)|> # aggregate story count by week for each candidate
  summarise(matched_stores = sum(matched_stories, na.rm = TRUE),
            all_candidate_stories = sum(all_candidate_stories, na.rm = TRUE),
            # calculate percentage of stories (online news) mentioned for each candidate
            pct_stories = (matched_stores/all_candidate_stories)*100,
            .groups = 'drop')|> 
  arrange(desc(pct_stories))|> # arrange pct_stores in descending order
  select(name, week, pct_stories)
```

We then determine the candidates with the highest and second-highest percentage of online news mentions each week, focusing particularly on Joe Biden, as he consistently ranks highest. From here, we can compute how many times higher Joe Biden's mentions are compared to the second highest candidate, which will be used for Figure 1.

```{r}
# Need to see who has the second highest pct_stories after Joe Biden

online_rank <- online_news|>
  group_by(week)|> # aggregate the data by week
  arrange(week, desc(pct_stories))|> # arrange by week and by pct_stories in descending order
  mutate(rank = rank(-pct_stories, ties.method = "first"))|> # rank by pct_stories for each candidate
  filter(rank %in% c(1, 2))|> # select the candidates with the top two highest pct_stories
  summarise(biden_pct_stories = first(pct_stories), # Joe Biden's pct_stories
            second_highest_pct_stories = last(pct_stories), # second highest pct_stories
            # ratio of Biden's pct_stories/ second rank's pct_stories
            times_pct_stories = (biden_pct_stories/second_highest_pct_stories),
            .groups = 'drop') 
```

### Data Wrangling: Cable News

We then proceed to process `cable_data` to compute the weekly percentage of cable news mentions for each candidate starting from 2019-05-01. 

```{r}
cable_news <- cable_data|>
  mutate(date = ymd(date))|> # convert date into date format
  filter(date >= "2019-05-01")|> # select data starting from 2019-05-01
  mutate(week = floor_date(date, unit = "week", week_start = 7))|> # week starts from Sunday
  #arrange()|>
  group_by(name, week)|> # aggregate clip count by week for each candidate
  summarise(matched_clips = sum(matched_clips, na.rm = TRUE),
            all_candidate_clips = sum(all_candidate_clips, na.rm = TRUE),
            total_clips = sum(total_clips, na.rm = TRUE),
            # calculte percentage of clips (cable news) mentioned for each candidate
            pct_clips = (matched_clips/all_candidate_clips)*100,
            .groups = 'drop')|>
  arrange(desc(pct_clips))|> # arrange pct_clips in descening order
  select(name, week, pct_clips)
```

Similar to the online news data, we determine the candidates with the highest and second-highest percentages of cable news mentions each week, again focusing on Joe Biden when calculating how much more frequently he is mentioned compared to the second-highest candidate. This calculation will also contribute to the visualisations for Figure 1.

```{r}
# Need to see who has the second highest pct_clips after Joe Biden

cable_rank <- cable_news|>
  group_by(week)|> # aggregate the data by week
  arrange(week, desc(pct_clips))|> # arrange by week and by pct_clips in descening order
  mutate(rank = rank(-pct_clips, ties.method = "first"))|> # rank the candidate by pct_clips 
  filter(rank %in% c(1, 2))|> # show only candidates with the top two highest pct_clips 
  summarise(biden_pct_clips = first(pct_clips),
            second_highest_pct_clips = last(pct_clips),
            # ratios of Biden's pct_clips/ second rank's pct_clips
            times_pct_clips = (biden_pct_clips/second_highest_pct_clips),
            .groups = 'drop')
```

#### Combine online_rank and cable_rank

After computing the weekly mentions ratios for both online and cable news, we combine these results to create a comprehensive view of Biden's media coverage relative to the second-ranking candidate across both platforms

After calculating the weekly mention ratios for both online and cable news, we combine these results to create a comprehensive view of Joe Biden's media coverage relative to the second-ranking candidate across both platforms. This step involves joining the datasets by week and selecting the relevant columns to form a consolidated table of ratios.

```{r}
ratio_table <- online_rank|>
  left_join(cable_rank, by = "week")|>
  select(week, times_pct_stories, times_pct_clips)
```

## Figure 1. Biden's Ukraine-related media bump is fading

### Description: Biden's share of media mentions on each medium relative to the next most-mentioned candidate each week from May 2019

```{r}
figure1 <- ratio_table|>
  ggplot(aes(x=week))+
  geom_line(aes(y=times_pct_stories, color = "Online News"), size = 1.2)+
  geom_line(aes(y=times_pct_clips, color = "Cable News"), size = 1.2)+
  labs(title = "Biden's share of media mentions on each medium relative to \nthe next most-mentioned candidate each week (2019)",
       x = "Month",
       y = "News Coverage")+
  scale_color_manual(values = c("Online News" = "darkorange2",
                                "Cable News" = "cadetblue3"))

ggplotly(figure1)
```

## Table 1. Tulsi Gabbard was mentioned more on cable last week

```{r}
# on the week of 10/13/19
# on the week of 10/20/19

# Calculate the difference in cable news for each candidate
cable_diff <- cable_news|>
  filter(week %in% c("2019-10-13",  "2019-10-20"))|> # as stated in the original Table
  mutate(pct_clips = round(pct_clips, 1))|> # round the data to one decimal place
  select(name, week, pct_clips)|>
  # convert data frame for calculating the difference
  pivot_wider(names_from = "week", values_from = "pct_clips")|> 
  mutate(diff = `2019-10-20` - `2019-10-13`) # calculate the difference in cable news mentioned

# Calculate the difference in online news for each candidate
online_diff <- online_news|>
  filter(week %in% c("2019-10-13",  "2019-10-20"))|> # as stated in the original Table 1
  mutate(pct_stories = round(pct_stories, 1))|> # round the data to one decimal place
  select(name, week, pct_stories)|>
  # convert data frame for calculating the difference
  pivot_wider(names_from = week, values_from = pct_stories)|>
  mutate(diff = `2019-10-20` - `2019-10-13`) # calculate difference in online news mentioned
```

```{r}
# Combine the two datasets for the cable and the online news diff for each candidate
table1 <- cable_diff|>
  full_join(online_diff, by = "name")|> # join the cable difference and online difference datasets
  rename("Candidate" = name,            # rename the column names
         "10/13/19 Cable (%)" = `2019-10-13.x`,
         "10/20/19 Cable (%)" = `2019-10-20.x`,
         "Diff Cable (%)" = `diff.x`,
         "10/13/19 Online (%)" = `2019-10-13.y`,
         "10/20/19 Online (%)" = `2019-10-20.y`,
         "Diff Online (%)" = `diff.y`)

kable(table1)
```
