---
title: "individaul_project"
output: html_document
date: "2025-01-20"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Article: The Media Frenzy Around Biden Is Fading

## Install Required Packages and Load Libraries
```{r}
#install.packages("dplyr")
#install.packages("lubridate")
#install.packages("ggplot2")
#install.packages("tidyr")
#install.packages("plotly")
#install.packages("knitr")

library(lubridate)
library(dplyr)
library(ggplot2)
library(tidyr)
library(plotly)
library(knitr)
```

## Import Data
```{r}
online_data <- read.csv("../data/online_weekly.csv")
cable_data <- read.csv("../data/cable_weekly.csv")
```
## Data Wrangling: online data - all candidate
```{r}
online_news <- online_data|>
  mutate(date = ymd(date))|> # convert date into ymd format
  filter(date >= "2019-05-01")|> # select data starting from 2019-05-01 as shown in the original figure 
  mutate(week = floor_date(date, unit = "week", week_start = 7))|> # week starts from Sunday
  arrange()|>
  group_by(name, week)|> # aggregate story count by week for each candidate
  summarise(matched_stores = sum(matched_stories, na.rm = TRUE),
            all_candidate_stories = sum(all_candidate_stories, na.rm = TRUE),
            pct_stories = (matched_stores/all_candidate_stories)*100)|>
  arrange(desc(pct_stories))|> # arrange pct_stores in descending order
  select(name, week, pct_stories)

head(online_news, 20)
```
```{r}
# Need to see who has the second highest pct_stories after Joe Biden

online_rank <- online_news|>
  group_by(week)|> # aggregate the data by week
  arrange(week, desc(pct_stories))|> # arrange the data by week and by pct_stories in descending order
  mutate(rank = rank(-pct_stories, ties.method = "first"))|> # rank by pct_stories for each candidate
  filter(rank %in% c(1, 2))|> # select the candidates with the top two highest pct_stories
  summarise(biden_pct_stories = first(pct_stories), # Joe Biden's pct_stories
            second_highest_pct_stories = last(pct_stories), # Candidate with the second highest pct_stories
            times_pct_stories = biden_pct_stories/second_highest_pct_stories) # ratio of Biden's pct_stories/ second rank's
```

## Data Wrangling: cable data - all candidates
```{r}
cable_news <- cable_data|>
  mutate(date = ymd(date))|> # convert date into date format
  filter(date >= "2019-05-01")|> # select data starting from 2019-05-01
  mutate(week = floor_date(date, unit = "week", week_start = 7))|> # week starts from Sunday
  #arrange()|>
  group_by(name, week)|> # aggregate clip count by week for each candidate
  summarise(matched_clips = sum(matched_clips, na.rm = TRUE),
            all_candidate_clips = sum(all_candidate_clips, na.rm = TRUE),
            total_clips = sum(total_clips, na.rm = TRUE),
            pct_clips = (matched_clips/all_candidate_clips)*100)|>
  arrange(desc(pct_clips))|> # arrange pct_clips in descening order
  select(name, week, pct_clips)
  
  
head(cable_news, 20)
```

```{r}
# Need to see who has the second highest pct_clips after Joe Biden

cable_rank <- cable_news|>
  group_by(week)|> # aggregate the data by week
  arrange(week, desc(pct_clips))|> #arrange the data by week and the clip count in descening order
  mutate(rank = rank(-pct_clips, ties.method = "first"))|> # rank the candidate by pct_clips 
  filter(rank %in% c(1, 2))|> # show only candidates with the top two highest pct_clips 
  summarise(biden_pct_clips = first(pct_clips),
            second_highest_pct_clips = last(pct_clips),
            times_pct_clips = biden_pct_clips/second_highest_pct_clips)
```

## Combine online_rank and cable_rank
```{r}
ratio_table <- online_rank|>
  left_join(cable_rank, by = "week")|>
  select(week, times_pct_stories, times_pct_clips)
```

## Figure 1. Biden's Ukraine-related media bump is fading
### Description: Biden's share of media mentions on each medium relative to the next most-mentioned candidate each week from May 2019
```{r}
figure1 <- ratio_table|>
  ggplot(aes(x=week))+
  geom_line(aes(y=times_pct_stories, color = "Online News"), size = 1.2)+
  geom_line(aes(y=times_pct_clips, color = "Cable News"), size = 1.2)+
  labs(title = "",
       x = "Month",
       y = "News Coverage")+
  scale_color_manual(values = c("Online News" = "darkorange2",
                                "Cable News" = "cadetblue3"))

ggplotly(figure1)
```

## Table 1. Tulsi Gabbard was mentioned more on cable last week
```{r}
# on the week of 10/13/19
# on the week of 10/20/19

# Calculate the difference in cable news for each candidate
cable_diff <- cable_news|>
  filter(week %in% c("2019-10-13",  "2019-10-20"))|> # select 2019-10-13 and 2019-10-20 as the original Table
  mutate(pct_clips = round(pct_clips, 1))|> # round the data to two decimal places
  select(name, week, pct_clips)|>
  pivot_wider(names_from = "week", values_from = "pct_clips")|> # convert data frame for calculating the difference
  mutate(diff = `2019-10-20` - `2019-10-13`) # calculate the difference

# Calculate the difference in online news for each candidate
online_diff <- online_news|>
  filter(week %in% c("2019-10-13",  "2019-10-20"))|>
  mutate(pct_stories = round(pct_stories, 1))|>
  select(name, week, pct_stories)|>
  pivot_wider(names_from = week, values_from = pct_stories)|>
  mutate(diff = `2019-10-20` - `2019-10-13`)
```

```{r}
# Combine the two datasets for the cable and the online news diff for each candidate
table1 <- cable_diff|>
  full_join(online_diff, by = "name")|> # join the cable difference and online difference datasets
  rename("CANDIDATE" = name,            # rename the column names
         "10/13/19 CABLE" = `2019-10-13.x`,
         "10/20/19 CABLE" = `2019-10-20.x`,
         "DIFF CABLE" = `diff.x`,
         "10/13/19 ONLINE" = `2019-10-13.y`,
         "10/20/19 ONLINE" = `2019-10-20.y`,
         "DIFF ONLINE" = `diff.y`)

kable(table1)
```

